High geomagnetic field intensity recorded by anorthosite xenoliths requires a strongly powered late Mesoproterozoic geodynamo

### 2022-04-04 PNAS_manuscript/PNAS_manuscript.tex

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to provide constraints on the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, likely necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained ca. 565 Ma during the Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and ca. 370 Ma during the Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a, Hawkins2021a}, potentially indicating unusual periods of core dynamo activity at those times. The Ediacaran data have been interpreted to indicate that there was a progressively decaying field up to that time that was soon followed by initial crystallization of the inner core \cite{Bono2019a}. Sparse data in the Proterozoic Era (2500 to 539 Ma) and the reality of high variability in Phanerozoic Era records (539 to 0 Ma) necessitate additional data to evaluate interpreted trends.

@@

Determinations of the absolute value of ancient geomagnetic field strength rely on igneous rocks that acquire thermal remanent magnetizations as they cool. These magnetizations need to be unmodified by subsequent heating or chemical alteration in order to maintain a record of the ancient geomagnetic field from the time of cooling. Intracontinental magmatic events are an important target for determination of ancient paleointensity as they can be well-preserved within continental interiors. This interior position results in them typically being distant from tectonic events along continental margins that can drive alteration through heat and fluids. However, intraplate magmatism associated with large igneous provinces is typically of geologically short duration with the bulk of magmatic products emplaced within 1 Myr or less \cite{Kasbohm2021a}. The Midcontinent Rift (Fig. \ref{fig:Geologic_map}) is an exception as it is a large igneous province where magmatism lasted $\sim$25 Myr from \texit{ca.} 1109 Ma to 1084 Ma during which there were pulsed intervals of more rapid activity \cite{Swanson-Hysell2021a}. Additionally, extension ceased in the Midcontinent Rift prior to lithospheric separation, preserving volcanic, intrusive, and sedimentary rocks of the rift within the continental interior far from the continental margin and subsequent orogenesis. As a result, rocks of the rift have unusually simple paleomagnetic behavior for their greater than one billion-year-old age and paleomagnetic data from rift rocks forms a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. The duration of magmatic activity within the Midcontinent Rift is longer than the entire 20.4 Myr long Neogene Period such that it enables an extended well-preserved window into the intensity of Earth's magnetic field in the late Mesoproterozoic.

Despite the excellent preservation of the rocks, non-ideal paleointensity behaviors have challenged the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a}. The most trusted type of paleointensity estimate is that obtained through experiments in which the primary natural thermal remanent magnetization is progressively replaced by a laboratory magnetization that is imparted in a known field with internal consistency checks (such as in IZZI-style Thellier experiments;  \citealp{Yu2004a}). In such Thellier paleointensity experiments, one typical departure from ideal behavior due to the presence of nonuniformly magnetized grains (with either multidomain \cite{Dunlop2001a} or vortex states \cite{Tauxe2020a}) is sagging or double-slopes as visualized in Arai plots that show thermal remanent magnetization (TRM) acquired versus natural remanent magnetization (NRM) lost. For such data, distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope. Typically, such non-ideal behavior would result in a higher paleointensity estimate from the steeper-sloped low-temperature portion of the experiment and a lower paleointensity estimate from the high-temperature portion. For example, in data from the Midcontinent Rift, ref. \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Such non-ideal results were rejected by ref. \citealp{Sprain2018a} who applied stricter paleointensity selection criteria, but as a result had few accepted sites.

In this study, we target the high-purity anorthosite xenoliths of the Beaver River diabase in the Midcontinent Rift. Anorthosites are cumulate rocks composed dominantly of the mineral plagioclase. They are attractive targets for paleomagnetic study as plagioclase crystals can enshroud magnetic inclusions from alteration. In addition, the weathering of the plagioclase crystals do not resulting in the formation of secondary iron oxides, whereas other minerals such as olivine and pyroxene are more likely to alter and influence the magnetic mineralogy. The anorthosite xenoliths targeted in this study were brought to the near surface in magma that formed hypabyssal (shallowly emplaced) intrusions of the Beaver River diabase \cite{Zhang2021b}. They likely cooled from tholeiitic magma temperatures and acquired their magnetizations in conjunction with the host diabase at a paleolatitude of 22 $\pm$ 2\textdegree (calculated from the paleomagnetic pole of the coeval Portage Lake Volcanics; \citealp{Swanson-Hysell2019a}). The details regarding the thermal history of these anorthosite xenoliths are discussed in \cite{Zhang2021b}. Paleointensity experiments on the anorthosite xenoliths have a high success rate, yielding consistent specimen- and site-level paleointensity results. The anorthosite xenoliths show low anisotropy of thermal remanent magnetization (TRM) acquisition and can acquire TRM linearly within the range of relevant field strengths. Magnetic imaging shows that the anorthosite specimens have dominant magnetic carriers within and interstitial to plagioclase crystals without strong preferred orientations. These single-slope, high-quality paleointensity data, together with other high-quality paleointensity records during the protracted Midcontinent Rift magmatic activity, require that there was a strong late Mesoproterozoic geodynamo.

\section*{Background on anorthosite xenoliths of the Beaver River diabase}

While magmatic activity within the Midcontinent Rift was protracted, there were intervals of particularly rapid volcanism and voluminous emplacement of intrusions \cite{Swanson-Hysell2021a}. The ca. 1092 Ma Beaver Bay Complex in northern Minnesota punctuates one such period of magmatism during the main stage of Midcontinent Rift activity. The magma that formed the 1091.7$\pm$0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. These anorthosite xenoliths are plagioclase cumulates that formed comagmatically with the host diabase in the lower crust---an interpretation confirmed by U-Pb zircon geochronology \cite{Zhang2021b}.

Thermal modeling results and paleomagnetic directional data show the anorthosite xenoliths to have acquired thermal remanent magnetizations while cooling with the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetization data show the anorthosite xenoliths to have dominantly single-component magnetization that often unblocks sharply within temperature ranges between 500\textdegree C and 580\textdegree C, consistent with magnetization held by low-titanium titanomagnetite \cite{Zhang2021b}.

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=13 cm]{Petro_QDM_full.png}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F,G,H) of anorthosite from the Beaver River anorthosite xenolith in the Silver Bay region from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column;  \citealp{Zhang2021b}) and a distinct anorthosite from the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture of crystals that are largely free of large opaque inclusions. In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [001] axis of the plagioclase. Maps of the vertical component of magnetic field (B$_z$) developed with a quantum diamond microscope (QDM) show relatively weak magnetic sources within plagioclase crystals in the Beaver River anorthosite (in E) relative to the strongly magnetic large oxide needles within Duluth Complex plagioclase (in F). The B$_z$ color scale is an order of magnitude greater in the Duluth Complex anorthosite maps (F, H) than the maps for the Beaver River anorthosite (E, G). (E to G) and (F to H) show experiments performed on both samples where we apply a first field along the y axis of the field of view and then apply a second field of 300 mT orthogonal to the first field direction. The remanent magnetizations acquired by both anorthosites were mapped after the application of each field. The magnetic images show that remanent magnetizations of Beaver River anorthosite (i.e. the individual dipoles visible with paired red +B$_z$ and blue -B$_z$ lobes) align well with the first applied field direction (E) and then rotated to align with the second applied field direction indicating minimal anisotropic behavior (G). In contrast, the remanence magnetizations carried by the magnetic needles in plagioclase 1 of the Duluth anorthosite xenolith align well with the first applied field but those in plagioclase 2 acquired an oblique remanence direction with respect to the field direction (F). After the application of an orthogonal 300 mT external field, magnetization of those needles in plagioclase 1 did not change direction due to strong shape anisotropy whereas the magnetization of needles in plagioclase 2 flipped, but with the acquired remanence remaining oblique to the field direction (H). This experiment illustrates that the Beaver River anorthosite xenolith has much lower magnetic anisotropy than the Duluth anorthosite xenolith. The unit of the axes in the QDM maps are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

@@

The dominantly monomineralic anorthosite xenoliths within the Beaver River diabase often have granoblastic texture characterized by equigranular crystals with weakly developed petrofabrics (Fig. \ref{fig:Petro_QDM}A). Plagioclase crystals within coarse-grained intrusions often contain abundant elongate Fe-Ti oxide inclusions that are visible through optical microscopy \cite{Feinberg2006a, Wenk2011a, Ageeva2016a}. The Beaver River anorthosite xenoliths lack such large (10s of $\mu$m) oxide inclusions such that oxides are not visible within the plagioclase crystals through optical microscopy (Fig. \ref{fig:Petro_QDM}C). However, magnetic imaging using a quantum diamond microscope (QDM) shows that there are magnetic remanence carriers within the plagioclase crystals (Fig. \ref{fig:Petro_QDM}E, G). The sources of these magnetic remanence carriers are likely to be Fe-Ti oxides that exsolved within plagioclase crystals above the Curie temperature of magnetite \cite{Zhang2021b, Bian2021a}. To highlight these distinct aspects of the Beaver River anorthosite xenoliths, we present petrographic and magnetic imaging data from the Beaver River anorthosite as well as a sample of an anorthosite from the Duluth Complex Anorthositic Series---an older intrusive complex within the Midcontinent Rift that was not targeted for paleointensity experiments in this study (Fig. \ref{fig:Petro_QDM}). In contrast to the Beaver River anorthosite, the plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display pronounced igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles within the Duluth Complex plagioclase grains that are typically aligned with the [001] axes of the crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that these magnetite needles have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}F, H). As a result of this shape anisotropy, magnetization is acquired at angles highly oblique to applied fields (Figure \ref{fig:Petro_QDM}F). In an experiment where a second orthogonal field was applied, the first isothermal remanent magnetizations (IRM) of a set of magnetite needles within a plagioclase grain either fail to rotate or flip by 180\textdegree. When they flip, they remain in a direction that is oblique to the applied field direction dictated by shape anisotropy (Fig. \ref{fig:Petro_QDM}F, H). These experiments enable novel visualization of the grain-scale magnetic anisotropy of elongated exsolved (titano)magnetite inclusions within plagioclase that leads to magnetic anisotropy observed at the bulk sample scale \cite{Selkin2000a, Feinberg2006a}. In contrast, the remanent magnetizations of the ferromagnetic grains imaged in the Beaver River anorthosite xenoliths, targeted for paleointensity in this study, align with the applied field directions indicating minimal remanence anisotropy  (Fig. \ref{fig:Petro_QDM}E, G). The relative lack of petrofabrics and minimal grain-scale magnetic anisotropy make the Beaver River anorthosite xenoliths a particularly compelling target for paleointensity experiments.

@@

Following IZZI-style paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria (see Materials and Methods section). 7 anorthosite sites and no diabase sites have specimen results that pass these selection criteria. Example NRM/TRM (Arai) plots are shown in Figure \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level mean paleointensity values are plotted in Figure \ref{fig:PINT_cooling_corrected} (and provided in Table S1) where each site is an individual anorthosite xenolith. The paleointensity quality index (Q$_{PI}$; \citealp{Biggin2014a}) for the anorthosite xenolith sites are all 5 or 6 (Table S2). The cooling rate-corrected absolute paleointensity estimates from the anorthosite xenoliths have a mean of 38.28$\pm$11.92 $\mu$T. The site-mean virtual dipole moment is $\sim$82 ZAm$^2$ ($10^{21} Am^2$) ca. 1092 Ma. All measurement-level paleointensity experiment data are available within the MagIC database (\url{https://earthref.org/MagIC/19462/8d3c2258-11ae-4830-b99f-3f6b02eceb7e}; \textit{this private link is provided for the purpose of review; the url will be updated when a doi is generated for this manuscript}).

\begin{figure*}
\noindent\includegraphics[width=17.8cm]{IZZI_examples.pdf}
\centering
\caption{\footnotesize{Example results of paleointensity experiments are displayed on Arai plots and zero-field heating results are shown on inset orthogonal plots (Zijderveld plots) for anorthosite and diabase specimens. Red (blue) circles indicate zero-field/in-field (in-field/zero-field) steps `ZI’ (`IZ’). Triangles mark partial thermal remanent magnetization (pTRM) checks. Blue and red squares in the Zijderveld plots are X–Y and X–Z projections, respectively, of the NRMs in specimen coordinates. Dashed red and blue lines show bounding regions calculated for determine the \textit{SCAT} statistic \cite{Shaar2013a}. Plots on the top row show successful specimen paleointensity results with straight, single-slope behaviors that pass our selection criteria. The green lines represent fits for the dominant single-slope component that passes the acceptance criteria and gives an estimate of the ancient field strength (B$_{anc}$). The plots for anorthosite specimens AX1-1a and diabase BD5-1b on the bottom row show non-ideal sagging behaviour that fails our acceptance criteria. Specimen BD2-5a is an example where the data appear linear with distinct slopes in the low and high temperature ranges such that it could pass less restrictive selection criteria, particularly if a narrower temperature range was used for the experiment. Data analysis and visualization was conducted using PmagPy \cite{Tauxe2016a}}}
\label{fig:IZZI_examples}
\end{figure*}

Typical paleointensity experimental data of the anorthosite specimens have straight, single-slope NRM/TRM plots and the accepted fractions of temperature steps span over the origin-trending, primary remanence components (Fig. \ref{fig:IZZI_examples}). We accept specimen- and site-level absolute paleointensity results from those anorthosite xenoliths that pass the selection criteria. Other anorthosite xenoliths and diabase specimens failed the selection criteria largely because of double-slope or sagging behavior (fail FRAC selection; see Materials and Methods section), poor pTRM checks, and sometimes zigzagging behaviors superimposed on top of sagging behavior (fail SCAT, DANG selection; see Materials and Methods section; Fig. \ref{fig:IZZI_examples}). A 20 mT alternating field treatment after in-field heating steps was applied to some specimens, but it did not result in significant changes in the experimental results for the anorthosite xenolith or diabase specimens (Fig. \ref{fig:PINT_cooling_corrected}; Supporting Information).

@@

In addition to estimating paleointensity values by introducing a set of selection criteria to filter our experiment results, we apply an independent method from ref. \citealp{Cych2021a} with all experimental data regardless of their NRM/TRM plot statistics to perform bias corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on an assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Paterson2011a}. The posterior paleointensity distributions from these sites with high-quality specimen-level data are in agreement with the site-level averages developed using the selection criteria approach (Fig. \ref{fig:PINT_cooling_corrected}; Fig. S1). Overall, the high-quality results from the anorthosite xenoliths of the Beaver River diabase indicate that the anorthosite xenoliths record a high geomagnetic field ca. 1092 Ma.

>>>>

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to provide constraints on the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, likely necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained ca. 565 Ma during the Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and ca. 370 Ma during the Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a, Hawkins2021a}, potentially indicating unusual periods of core dynamo activity at those times. The Ediacaran data have been interpreted to indicate that there was a progressively decaying field up to that time that was soon followed by initial crystallization of the inner core \cite{Bono2019a}. Sparse paleointensity data in the Proterozoic Era (2500 to 539 Ma) and the reality of high variability in Phanerozoic Era records (539 to 0 Ma) necessitate additional data to evaluate interpreted trends.

@@

Determinations of the absolute value of ancient geomagnetic field strength rely on igneous rocks that acquire thermal remanent magnetizations as they cool. These magnetizations need to be unmodified by subsequent heating or chemical alteration in order to maintain the record of the ancient geomagnetic field from the time of cooling. Intracontinental magmatic events are an important target for determination of ancient paleointensity as they can be well-preserved within continental interiors. This interior position results in them typically being distant from tectonic events along continental margins that can drive alteration through heat and fluid flow. However, intraplate magmatism associated with large igneous provinces is typically of geologically short duration with the bulk of magmatic products emplaced within 1 Myr or less \cite{Kasbohm2021a}. The Midcontinent Rift (Fig. \ref{fig:Geologic_map}) is an exception as it is a large igneous province where magmatism lasted $\sim$25 Myr from \texit{ca.} 1109 Ma to 1084 Ma during which there were pulsed intervals of more rapid activity \cite{Swanson-Hysell2021a}. Additionally, extension ceased in the Midcontinent Rift prior to lithospheric separation, preserving volcanic, intrusive, and sedimentary rocks of the rift within the continental interior far from the continental margin and subsequent orogenesis. As a result, rocks of the rift have unusually simple paleomagnetic behavior for their greater than one billion-year-old age and paleomagnetic data from rift rocks form a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. The duration of magmatic activity within the Midcontinent Rift is longer than the entire 20.4 Myr long Neogene Period such that it provides an extended well-preserved window into the intensity of Earth's magnetic field in the late Mesoproterozoic.

Despite the excellent preservation of the rocks, non-ideal paleointensity behaviors have thwarted the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a}. The most trusted type of paleointensity estimate is that obtained through experiments in which the primary natural remanent magnetization (NRM) is progressively replaced by a laboratory magnetization that is imparted in a known field with internal consistency checks (such as in IZZI-style Thellier experiments;  \citealp{Yu2004a}). In such Thellier paleointensity experiments, one typical departure from ideal behavior due to the presence of nonuniformly magnetized grains (with either multidomain \cite{Dunlop2001a} or vortex states \cite{Tauxe2020a}) is sagging or double-slopes as visualized in Arai plots that show thermal remanent magnetization (TRM) acquired versus NRM lost. For such data, distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope. Typically, such non-ideal behavior would result in a higher paleointensity estimate from the steeper-sloped low-temperature portion of the experiment and a lower paleointensity estimate from the high-temperature portion. For example, in data from the Midcontinent Rift, ref. \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Such non-ideal results were rejected by ref. \citealp{Sprain2018a} who applied stricter paleointensity selection criteria, but as a result had few accepted sites.

In this study, we target the high-purity anorthosite xenoliths of the Beaver River diabase in the Midcontinent Rift. While magmatic activity within the Midcontinent Rift was protracted, there were intervals of particularly rapid volcanism and voluminous emplacement of intrusions \cite{Swanson-Hysell2021a}. The ca. 1092 Ma Beaver Bay Complex in northern Minnesota punctuates one such period of magmatism during the main stage of Midcontinent Rift activity. The magma that formed the 1091.7$\pm$0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. These anorthosite xenoliths are plagioclase cumulates that formed comagmatically with the host diabase in the lower crust---an interpretation confirmed by U-Pb zircon geochronology \cite{Zhang2021b}. Anorthosites are cumulate rocks composed dominantly of the mineral plagioclase. They are attractive targets for paleomagnetic study as plagioclase crystals can protect magnetic inclusions from alteration. In addition, the weathering of the plagioclase crystals do not resulting in the formation of secondary iron oxides, whereas other minerals such as olivine and pyroxene are more likely to alter and influence the magnetic mineralogy. The anorthosite xenoliths targeted in this study were brought to the near surface in magma that formed hypabyssal (shallowly emplaced) intrusions of the Beaver River diabase \cite{Zhang2021b}. They were likely heated to tholeiitic magma temperatures and partially recrystallized \cite{Zhang2021b}. Fe-Ti oxides within the plagioclase exsolved above magnetite Curie temperature ($\sim$580\textdegree C; \citealp{Bian2021a}) and subsequently cooled and acquired thermal remanent magnetizations in conjunction with the host diabase at a paleolatitude of 22 $\pm$ 2\textdegree (calculated from the paleomagnetic pole of the coeval Portage Lake Volcanics; \citealp{Swanson-Hysell2019a}). The details regarding the thermal history of these anorthosite xenoliths are discussed in \cite{Zhang2021b}. Paleointensity experiments on the anorthosite xenoliths have a high success rate, yielding consistent specimen- and site-level paleointensity results. The anorthosite xenoliths show low anisotropy of thermal remanent magnetization (TRM) acquisition and can acquire TRM linearly within the range of relevant field strengths. Magnetic imaging shows that the anorthosite specimens have dominant magnetic carriers within and interstitial to plagioclase crystals without strong preferred orientations. Thermal modeling results and paleomagnetic directional data show that the anorthosite xenoliths to have acquired thermal remanent magnetizations while cooling with the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetization data show the anorthosite xenoliths to have dominantly single-component magnetizations that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C, consistent with remanence being held by low-titanium titanomagnetite \cite{Zhang2021b}.

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=13 cm]{Petro_QDM_full.png}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F,G,H) of an anorthosite sample from the Beaver River anorthosite xenolith in the Silver Bay region from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column;  \citealp{Zhang2021b}) and from a distinct anorthosite within the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture of crystals that are largely free of large opaque inclusions. In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [001] axis of the plagioclase. Maps of the vertical component of magnetic field (B$_z$) developed with a quantum diamond microscope (QDM) show relatively weak magnetic sources within plagioclase crystals in the Beaver River anorthosite (in E) relative to the strongly magnetic large oxide needles within Duluth Complex plagioclase (in F). The B$_z$ color scale is an order of magnitude greater in the Duluth Complex anorthosite maps (F, H) than the maps for the Beaver River anorthosite (E, G). Panels (E) to (G) and (F) to (H) show experiments performed on both samples where we apply a first field along the y axis of the field of view and then apply a second field of 300 mT orthogonal to the first field direction. The remanent magnetizations acquired by both anorthosites were mapped after the application of each field. The magnetic images show that remanent magnetizations of Beaver River anorthosite (i.e. the individual dipoles visible with paired red +B$_z$ and blue -B$_z$ lobes) align well with the first applied field direction (E) and then rotated to align with the second applied field direction indicating minimal anisotropic behavior (G). In contrast, the remanence magnetizations carried by the magnetic needles in plagioclase 1 of the Duluth anorthosite xenolith align well with the first applied field but those in plagioclase 2 acquired an oblique remanence direction with respect to the field direction (F). After the application of an orthogonal 300 mT external field, magnetization of those needles in plagioclase 1 did not change direction due to strong shape anisotropy whereas the magnetization of needles in plagioclase 2 flipped, but with the acquired remanence remaining oblique to the field direction (H). This experiment illustrates that the Beaver River anorthosite xenolith has much lower magnetic anisotropy than the Duluth anorthosite xenolith. The unit of the axes in the QDM maps are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

@@

The dominantly monomineralic anorthosite xenoliths within the Beaver River diabase often have granoblastic texture characterized by equigranular crystals with weakly developed petrofabrics (Fig. \ref{fig:Petro_QDM}A). Plagioclase crystals within coarse-grained intrusions often contain abundant elongate Fe-Ti oxide inclusions that are visible with optical microscopy \cite{Feinberg2006a, Wenk2011a, Ageeva2016a}. The Beaver River anorthosite xenoliths lack such large (10s of $\mu$m) oxide inclusions such that they are not visible within the plagioclase crystals using optical microscopy (Fig. \ref{fig:Petro_QDM}C). Magnetic imaging using a quantum diamond microscope (QDM), however, shows that there are magnetic remanence carriers within the plagioclase crystals (Fig. \ref{fig:Petro_QDM}E, G). The sources of these magnetic remanence carriers are likely to be Fe-Ti oxides that exsolved within plagioclase crystals above the Curie temperature of magnetite \cite{Zhang2021b, Bian2021a}. To highlight these distinctive aspects of the Beaver River anorthosite xenoliths, we present petrographic and magnetic imaging data from the Beaver River anorthosite as well as an anorthosite sample from the Duluth Complex Anorthositic Series---an older intrusive complex within the Midcontinent Rift that was not targeted for paleointensity experiments in this study (Fig. \ref{fig:Petro_QDM}). In contrast to the Beaver River anorthosite, the plagioclase crystals of the Duluth Complex anorthosites typically develop interlocking textures that display pronounced igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles within the Duluth Complex plagioclase grains that are typically aligned with the [001] axes of the crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that these magnetite needles have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}F, H). As a result of this shape anisotropy, remanent magnetizations are acquired at angles highly oblique to applied fields (Figure \ref{fig:Petro_QDM}F). In an experiment where a second orthogonal field was applied, the first isothermal remanent magnetizations (IRM) of a set of magnetite needles within a plagioclase grain either fail to rotate or flip by 180\textdegree. When they flip, they remain in a direction that is oblique to the applied field direction dictated by shape anisotropy (Fig. \ref{fig:Petro_QDM}F, H). These experiments enable novel visualization of the grain-scale magnetic anisotropy of elongated exsolved (titano)magnetite inclusions within plagioclase grains that leads to magnetic anisotropy observed at the bulk sample scale \cite{Selkin2000a, Feinberg2006a}. In contrast, the remanent magnetizations of the ferromagnetic grains imaged in the Beaver River anorthosite xenoliths, targeted for paleointensity in this study, align with the applied field directions indicating minimal remanence anisotropy  (Fig. \ref{fig:Petro_QDM}E, G). The relative lack of petrofabrics and minimal grain-scale magnetic anisotropy make the Beaver River anorthosite xenoliths a particularly compelling target for paleointensity experiments.

@@

Following IZZI-style paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria (see Materials and Methods section). Seven anorthosite sites and no diabase sites have specimen results that pass these selection criteria. Example NRM/TRM (Arai) plots are shown in Figure \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level mean paleointensity values are plotted in Figure \ref{fig:PINT_cooling_corrected} (and provided in Table S1) where each site represents an individual anorthosite xenolith. The paleointensity quality index (Q$_{PI}$; \citealp{Biggin2014a}) for the anorthosite xenolith sites are all 5 or 6 (Table S2). The cooling rate-corrected absolute paleointensity estimates from the anorthosite xenoliths have a mean of 38.28$\pm$11.92 $\mu$T. The site-mean virtual dipole moment is $\sim$82 ZAm$^2$ ($10^{21} Am^2$) at ca. 1092 Ma. All measurement-level paleointensity experiment data are available within the MagIC database (\url{https://earthref.org/MagIC/19462/8d3c2258-11ae-4830-b99f-3f6b02eceb7e}; \textit{this private link is provided for the purpose of review; the url will be updated when a doi is generated for this manuscript}).

\begin{figure*}
\noindent\includegraphics[width=17.8cm]{IZZI_examples.pdf}
\centering
\caption{\footnotesize{Example results of paleointensity experiments are displayed on Arai plots and zero-field heating results are shown as inset orthogonal plots (Zijderveld plots) for anorthosite and diabase specimens. Red (blue) circles indicate zero-field/in-field (in-field/zero-field) steps `ZI’ (`IZ’). Triangles mark partial thermal remanent magnetization (pTRM) checks. Blue and red squares in the Zijderveld plots are X–Y and X–Z projections, respectively, of the NRMs in specimen coordinates. Dashed red and blue lines show bounding regions calculated for determine the \textit{SCAT} statistic \cite{Shaar2013a}. Plots on the top row show successful specimen paleointensity results with straight, single-slope behaviors that pass our selection criteria. The green lines represent fits for the dominant single-slope component that passes the acceptance criteria and gives an estimate of the ancient field strength (B$_{anc}$). The plots for anorthosite specimens AX1-1a and diabase BD5-1b on the bottom row show non-ideal sagging behaviour that fails our acceptance criteria. Specimen BD2-5a is an example where the data appear linear with distinct slopes in the low and high temperature ranges such that it could pass less restrictive selection criteria, particularly if a narrower temperature range was used for the experiment. Data analysis and visualization was conducted using PmagPy \cite{Tauxe2016a}}}
\label{fig:IZZI_examples}
\end{figure*}

Typical paleointensity experimental data of the anorthosite specimens have straight, single-slope NRM/TRM plots and the accepted fractions of temperature steps span over the origin-trending, primary remanence components (Fig. \ref{fig:IZZI_examples}). We accept specimen- and site-level absolute paleointensity results from those anorthosite xenoliths that pass the selection criteria. Other anorthosite xenoliths and diabase specimens failed the selection criteria largely because of double-slope or sagging behavior (fail FRAC selection; see Materials and Methods section), poor pTRM checks, and zigzagging behaviors occasionally superimposed on top of sagging behavior (fail SCAT, DANG selection; see Materials and Methods section; Fig. \ref{fig:IZZI_examples}). A 20 mT alternating field treatment after in-field heating steps was applied to some specimens, but this treatment did not result in significant changes in the experimental results for the anorthosite xenolith or diabase specimens (Fig. \ref{fig:PINT_cooling_corrected}; Supporting Information).

@@

In addition to estimating paleointensity values by introducing a set of selection criteria to filter our experiment results, we applied an independent statistical method from ref. \citealp{Cych2021a} to all experimental data regardless of their NRM/TRM plot statistics to make a bias-corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on the assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Paterson2011a}. The posterior paleointensity distributions from these sites with high-quality specimen-level data are in agreement with the site-level averages developed using the selection criteria approach (Fig. \ref{fig:PINT_cooling_corrected}; Fig. S1). Overall, the high-quality results from the anorthosite xenoliths of the Beaver River diabase indicate that the anorthosite xenoliths record a high geomagnetic field at ca. 1092 Ma.

\subsection*{Rock magnetism}
\subsubsection*{Coercivity}
Additional rock magnetic data indicate that those anorthosite specimens which pass the paleointensity selection criteria contain dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection along with all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show that both the diabase and anorthosite contain (titano)magnetite as evidenced by the presence of the Verwey transition (Fig. S2;  \citealp{Verwey1939a, Feinberg2015a}). Anorthosite specimens from sites that yield successful paleointensity results have Verwey transition temperatures near 120 K as expected for stochiometric magnetite with minimal Ti contents \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have Verwey transitions that are trending toward lower temperatures (Fig. S2), indicating that their magnetic grains either have relatively higher Ti content or have been partially oxidized \cite{Ozdemir1993a}. Another difference is that samples which pass paleointensity selection have distinctly higher average median destructive field values than other anorthosite and diabase specimens (Fig. \ref{fig:coercivity}). Single-component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites having successful paleointensity results can contain magnetic grain populations with peak coercivities ranging from $\sim$44-144 mT with a median of 53 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase specimens tend to have lower peak coercivities ($\sim$30 mT). This result is consistent with an interpretation that magnetic grain populations with more multidomain-like behavior are responsible for the non-ideal paleointensity behaviors during experiments of such specimens \cite{Xu2004a}.
